% Acknowledgements

\pdfbookmark[1]{Acknowledgements}{Acknowledgements} % Bookmark name visible in a PDF viewer

\begingroup

\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter*{Acknowledgements} % Acknowledgements section text

This thesis is the result of 4 and a half years of research during which I have been supported by many helpful people.

First and foremost, I thank my supervisor Sergei Gorlatch how has given me the guidance and freedom which enabled my research.
I'm especially thankful for this support of my numerous and fruitful research visits at the University of Edinburgh.

I thank Christophe Dubach for our joint research collaboration with many rich and extensive discussions.
I'm especially grateful for this continuous enthusiasm for our collaborative work.

During my time in M{\"u}nster I had the pleasure of a friendly and supportive group of colleagues, which includes:
Sebastian Albers,
Frank Glinka,
Waldemar Gorus,
Michael Haidl,
Tim Humernbrum,
Julia Kaiser-Mariani,
Philipp Kegel
Dominik Meil{\"a}nder,
Mohammed Nsaif,
Alexander Ploss,
and
Ari Rasch.

I want to specifically thank Philipp Kegel for our numerous discussions and close collaboration on \SkelCL and Sebastian Albers and Michael Haidl for their work on our joint publications.

I also want to mention and thank two friends and fellow PhD students from M{\"u}nster Benjamin Risse and Sven Strothoff with whom I could share the day-to-day struggles of a PhD student.

During the work on this thesis I visited the University of Edinburgh numerous times and eventually started working there as a Research Associate.
During my visits I have met many friendly, interesting, and inspiring people -- too many to list all of them here.
I want to especially thank Thibaut Lutz which whom I collaborated closely during my first stay in Edinburgh and his supervisors Christian Fensch and Murray Cole.
I already mentioned Christophe Dubach which whom I started a fruitful research collaboration leading to the development of the code generation technique presented in the second half of this thesis.

I'm grateful for the financial support I received from the EU founded HPCEuropa2 program and the HiPEAC network of excellence which made my research visits to Edinburgh possible.

As I spend a large part of my time in M{\"u}nster teaching, supervising, and working with students I want to thank all the students participating in the 14 courses I co-supervised.
By teaching students computer science I gained many useful skills which have contributed to the quality of my research.

I particular want to mention and thank the students I have co-supervised during their Bachelor and Master theses.
These are:
Markus Blank-Burian,
Stefan Breuer,
Julian Buscher,
Matthias Buss,
Matthias Droste,
Malte Friese,
Tobias G{\"u}nnewig,
Wadim Hamm,
Kai Kientopf,
Lars Klein.
Sebastian Missbach,
Michael Olejnik,
Florian Quinkert,
Patrick Schiffler,
and
Jan-Gerd Tenberge.

% Family
Last, but not least I thank my parents and my brothers for their support, love, and interest in me and my research.
Especially, for their continuing engagement and enthusiasm about my research visits to universities and conferences around the world.
%the location of my next research trip.

\bigskip
\begin{flushright}
\emph{Michel Steuwer}\\
April 2015\\
Edinburgh
\end{flushright}


\endgroup

