% Chapter 5: Code generation using patterns

\chapter{Code generation using patterns}
\addtocontents{lof}{\protect\vspace{\beforebibskip}}%
\addtocontents{lot}{\protect\vspace{\beforebibskip}}%
\addtocontents{lol}{\protect\vspace{\beforebibskip}}%

\label{ch:fifth} % For referencing the chapter elsewhere, use \autoref{ch:name} 
\label{chapter:codeGeneration}

\lettrine[lines=3, loversize=0.1]{I}{n the previous two chapters} we discussed how regular parallel patterns, referred to as \emph{algorithmic skeletons}, help to simplify programming of modern parallel systems.
Parallel programming of multi-\GPU systems is considerably simplified without sacrificing performance as shown by the evaluation in \autoref{chapter:skelcl-evaluation}.

In this chapter, we address the second main challenge identified in \autoref{ch:introduction}: \emph{Performance portability}.
We will present a novel approach for generating efficient and hardware-specific code from compositions of high-level \emph{patterns}.
%algorithmic skeletons and similar 
This systematic, rule-based approach provides performance portability across different types of modern parallel processors. 
In the following \autoref{chapter:codeGeneration-evaluation} we will use a set of example applications to show that our approach generates code matching the performance of highly tuned implementations on \CPUs and \GPUs.

We will start this chapter by looking at optimizations in \OpenCL and how applying them changes the source code of applications.
We will show that these optimizations are often hardware-specific thus breaking performance portability: optimizations for one particular hardware architecture can lead to poor performance on other hardware architectures.
This will motivate the necessity, when aiming for performance portability, for generating optimized and specialized code from a pattern-based high-level representation.
We will discuss the performance benefits this code generation approach offers over a library-based approach like \SkelCL presented in \autoref{chapter:skelcl}.
Then we will give an overview of our approach and present it in more detail in the following sections.
\autoref{chapter:codeGeneration-evaluation} will present an evaluation of the approach using a set of application studies.

\newpage

\input{Chapters/Chapter5/motivation.tex}

\input{Chapters/Chapter5/overview.tex}

\input{Chapters/Chapter5/patterns.tex}

\input{Chapters/Chapter5/rules.tex}

\input{Chapters/Chapter5/generator.tex}

\section{Conclusion}
In this chapter we introduced a novel code generation technique for pattern-based programs.
We started the chapter with an investigation of the portability of optimizations in \OpenCL.
We concluded, that optimizations are not performance portable in \OpenCL and argued for a more systematic approach.
In the following we introduced a set of high-level algorithmic patterns -- similar to the algorithmic skeletons defined in \SkelCL in Part 2 -- but also low-level \OpenCL-specific patterns resembling the \OpenCL programming model.
We presented a formal system of rewrite rules which allow for systematically rewriting of pattern-based expressions, especially, for transforming programs expressed with the high-level algorithmic patterns into \OpenCL-specific programs expressed with the low-level patterns.
We showed the soundness of our approach by proving that the rewrite rules do not change the program semantics.
Finally, we presented an \OpenCL code generator which generates \OpenCL code for programs expressed with the low-level rules.

The rewrite rules encode different algorithmic as well as low-level optimization choices.
By applying the rules we derived expressions resembling manually optimized \OpenCL code written by Nvidia and performed an optimization to fuse two pattern, enabling the generation of an optimized implementation for the \emph{asum} benchmark, which we could not generate with \SkelCL.

In the next chapter, we will evaluate the performance of the systematically generated \OpenCL code.
We will also investigate and evaluate a prototype search tool which applies the rewrite rules automatically to find good implementations on three different hardware architectures for achieving performance portability.

